\stktokens{}: Enforcing Well-Bracketed Control Flow and Stack Encapsulation Using Linear Capabilities

### 2019-05-21 popl-2019-journal-v/proving-full-abstraction.tex
[POPLjv] Add description of readXCond

With the future region relation in place, we define the future world relation as follows:
\begin{mathpar}
  \inferrule{ \text{for $i \in \{\mathrm{heap},\mathrm{free},\mathrm{priv} \}$} \\ \exists m_i : \RegionName \fun \RegionName, \text{ injective}\ldotp \dom(W'.i) \supseteq \dom(m_i(W.i)) \wedge \forall r \in \dom(W.i)\ldotp W'.i(m_i(r)) \future W.i(r) }
            { W' \future W }
\end{mathpar}
The relation says that each of the three worlds must be an extension of the past world and for each of the existing regions that must be a future region.
% injective function
Note that the future world relation has a mapping function $m_i$ which allows us to change the naming of regions in future worlds.
The definition is a generalization of the standard definition where $m_i$ would be the identity\footnote{In \citet{Skorstengaard_reasoning_2017} the future region relation and the reasoning about the awkward example could have been simplified with this future world relation.}.

@@

The $\execCond{}$ intuitively says that an execute capability is safe when any capability that can be derived from it is safe as a program counter now and in the future.
We later define the $\lre$-relation which captures what it means for a word to be safe as a program counter, but for now it suffices to think of it as a program counter that causes an execution that does not break memory invariants.
An executable capability can have its range of authority shrunk or its current address changed which changes what it does and thus potentially whether it keeps memory invariants.
For this reason, the condition requires that any capability with a derived range of authority and a current address that range is safe to use for execution.
The $\execCond{}$ is quantified over all future worlds of the $\purePart{}$ of $W$. The $\purePart{}$ function strips the world of all spatial ownership.
This means that the capability cannot depend on linear capabilities, and when we define the logical relation, we even require the executable capability to not be linear.
It is not clear whether a linear executable capability would even be useful because there is no way to move it out of the $\pcreg$-register without crashing the execution.
This may sound like an ideal option for constructing something that can be executed once, however, most programs rely on loading other capabilities or seal sets for the program counter which is not possible.
\begin{align*}
  \execCond{W} &=
  \left\{ \npair{A} \middle|
    \begin{array}{l}
      \forall n' < n, W' \future \purePart{W}\ldotp \forall b',e'\ldotp \forall \aaddr \in [b',e'] \subseteq A \ldotp\\
      \quad \npair[n']{\stpair[.]{((\rx,\normal),\baddr',\eaddr',\aaddr)}{((\rx,\normal),\baddr',\eaddr',\aaddr)}} \in \lre(W')
    \end{array}
    \right\}
\end{align*}

% TODO Motivate the read execute condition (do this when description of code region done)
\[
  \xReadCond{W} = \left\{ \npair{A} \middle| 
    \begin{array}{l}
      \exists r \in \addressable{\normal, \pwheap}, \code \ldotp \\
      \qquad \pwheap(r) \nequal \codereg{\_,\_,\code,\gc}\wedge\\
      \qquad \dom(\code) \supseteq A 
    \end{array}
  \right\}
\]
% TODO Stack not executable, so no stack readx or stack exec cond.

\paragraph{The untrusted value relation}
\label{par:untrusted-val-rel}
\[
  \lrv(W) =
  \begin{array}[t]{l}
    \left\{ \npair{\stpair[.]{i}{i}} \;\middle|\; i \in \ints \right\}\cup \\
%
    \hspace{-2cm}\left\{ \npair{\left(\arraycolsep=0pt\array{l}\src{\sealed{\sigma,\vsc_S}},\\ \sealed{\sigma,\vsc_T} \endarray\right)} \;\middle| \;
    \begin{array}{l}
      (\isLinear{\src{\vsc_S}} \text{ iff } \isLinear{\vsc_T}) \wedge\\
      \exists r \in \dom(\pwheap), \sigrets,\sigcloss,\mscode \ldotp \pwheap(r) = (\pure,\_,H_\sigma) \tand \\
      \quad H_\sigma \; \sigma \nequal H^\mathrm{code,\square}_\sigma \; \sigrets \; \sigcloss \; \mscode \; \gc \; \sigma \tand \\
      \quad \npair[n']{\stpair[.]{\vsc_S}{\vsc_T}} \in H_\sigma \; \sigma \; \xi^{-1}(W) \text{ for all $n' < n$}\wedge\\
      \quad (\isLinear{\src{\vsc_S}} \Rightarrow \\
      \qquad\forall W' \future W, W_o, n' < n, \npair[n']{\stpair[.]{\vsc_S'}{\vsc'_T}} \in H_\sigma \; \sigma \; \xi^{-1}(W_o) \ldotp \\
      \qquad \quad \npair[n']{\src{\vsc_S},\src{\vsc_S'},\vsc_T,\vsc_T'} \in \lrexj(W'\oplus W_o)) \wedge \\
      \quad (\nonLinear{\src{\vsc_S}} \Rightarrow \\
      \qquad \forall W' \future \purePart{W}, W_o, n' < n, \npair[n']{\stpair[.]{\vsc_S'}{\vsc'_T}} \in H_\sigma \; \sigma \; \xi^{-1}(W_o) \ldotp \\
      \qquad \quad \npair[n']{\src{\vsc_S},\src{\vsc_S'},\vsc_T,\vsc_T'} \in \lrexj(W'\oplus W_o))

>>>>

With the future region relation in place, we define the future world relation as follows:
\begin{mathpar}
  \inferrule{ \text{for $i \in \{\mathrm{heap},\mathrm{free},\mathrm{priv} \}$} \\ \exists m_i : \RegionName \fun \RegionName, \text{ injective}\ldotp \dom(W'.i) \supseteq \dom(m_i(W.i)) \wedge \forall r \in \dom(W.i)\ldotp W'.i(m_i(r)) \future W.i(r) }
            { W' \future W }
\end{mathpar}
The relation says that each of the three worlds must be an extension of the past world and for each of the existing regions that must be a future region.
% injective function
Note that the future world relation has a mapping function $m_i$ which allows us to change the naming of regions in future worlds.
The definition is a generalization of the standard definition where $m_i$ would be the identity\footnote{In \citet{skorstengaard_reasoning_2017} the future region relation and the reasoning about the awkward example could have been simplified with this future world relation.}.

@@

The $\execCond{}$ intuitively says that an execute capability is safe when any capability that can be derived from it is safe as a program counter now and in the future.
We later define the $\lre$-relation which captures what it means for a word to be safe as a program counter, but for now it suffices to think of it as a program counter that causes an execution that does not break memory invariants.
An executable capability can have its range of authority shrunk or its current address changed which changes what it does and thus potentially whether it keeps memory invariants.
For this reason, the condition requires that any capability with a derived range of authority and a current address that range is safe to use for execution.
The $\execCond{}$ is quantified over all future worlds of the $\purePart{}$ of $W$. The $\purePart{}$ function strips the world of all spatial ownership.
This means that the capability cannot depend on linear capabilities, and when we define the logical relation, we even require the executable capability to not be linear.
It is not clear whether a linear executable capability would even be useful because there is no way to move it out of the $\pcreg$-register without crashing the execution.
This may sound like an ideal option for constructing something that can be executed once, however, most programs rely on loading other capabilities or seal sets for the program counter which is not possible.
  \execCond{W} &=
  \left\{ \npair{A} \middle|
      \forall n' < n, W' \future \purePart{W}\ldotp \forall b',e'\ldotp \forall \aaddr \in [b',e'] \subseteq A \ldotp\\
      \quad \npair[n']{\stpair[.]{((\rx,\normal),\baddr',\eaddr',\aaddr)}{((\rx,\normal),\baddr',\eaddr',\aaddr)}} \in \lre(W')
The $\readCond{}$ condition by itself allows many different regions and thus potentially many different memories.
However, when we have a read capability with execute permission, we know that the capability must point to a piece of this code memory.
For this reason, we define the $\xReadCond{}$ to capture the additional assumptions that we can make when a capability is executable.
  \xReadCond{W} = \left\{ \npair{A} \middle| 
      \exists r \in \addressable{\normal, \pwheap}, \code \ldotp \\
      \qquad \pwheap(r) \nequal \codereg{\_,\_,\code,\gc}\wedge\\
      \qquad \dom(\code) \supseteq A 
The $\xReadCond{}$ requires that the memory segment an executable capability has authority over is governed by a code region.
Note that we do not define $\execCond{}$ and $\xReadCond{}$ for the stack because the stack is not executable.
\paragraph{The untrusted value relation}
\label{par:untrusted-val-rel}
  \lrv(W) =
  \begin{array}[t]{l}
    \left\{ \npair{\stpair[.]{i}{i}} \;\middle|\; i \in \ints \right\}\cup \\
    \hspace{-2cm}\left\{ \npair{\left(\arraycolsep=0pt\array{l}\src{\sealed{\sigma,\vsc_S}},\\ \sealed{\sigma,\vsc_T} \endarray\right)} \;\middle| \;
      (\isLinear{\src{\vsc_S}} \text{ iff } \isLinear{\vsc_T}) \wedge\\
      \exists r \in \dom(\pwheap), \sigrets,\sigcloss,\mscode \ldotp \pwheap(r) = (\pure,\_,H_\sigma) \tand \\
      \quad H_\sigma \; \sigma \nequal H^\mathrm{code,\square}_\sigma \; \sigrets \; \sigcloss \; \mscode \; \gc \; \sigma \tand \\
      \quad \npair[n']{\stpair[.]{\vsc_S}{\vsc_T}} \in H_\sigma \; \sigma \; \xi^{-1}(W) \text{ for all $n' < n$}\wedge\\
      \quad (\isLinear{\src{\vsc_S}} \Rightarrow \\
      \qquad\forall W' \future W, W_o, n' < n, \npair[n']{\stpair[.]{\vsc_S'}{\vsc'_T}} \in H_\sigma \; \sigma \; \xi^{-1}(W_o) \ldotp \\
      \qquad \quad \npair[n']{\src{\vsc_S},\src{\vsc_S'},\vsc_T,\vsc_T'} \in \lrexj(W'\oplus W_o)) \wedge \\
      \quad (\nonLinear{\src{\vsc_S}} \Rightarrow \\
      \qquad \forall W' \future \purePart{W}, W_o, n' < n, \npair[n']{\stpair[.]{\vsc_S'}{\vsc'_T}} \in H_\sigma \; \sigma \; \xi^{-1}(W_o) \ldotp \\
      \qquad \quad \npair[n']{\src{\vsc_S},\src{\vsc_S'},\vsc_T,\vsc_T'} \in \lrexj(W'\oplus W_o))
